Statistical analysis of periodic data in neuroscience

### 2020-12-31 manuscript/manuscript.tex
Fixed cluster correction function

\title{Statistical analysis of periodic data in neuroscience}
\author{Daniel H. Baker}
\date{13/12/2020}

@@

Many experimental paradigms in neuroscience involve driving the nervous system with periodic sensory stimuli. Neural signals recorded with a variety of techniques will then include phase-locked oscillations at the stimulation frequency. The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components. However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded. An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components. Here we compare the performance of two multivariate extensions of the T-test, Hotelling's \(T^2\) and a variant called \(T^2_{circ}\). We develop a novel test of the assumptions of \(T^2_{circ}\) based on the condition index, and a heuristic for excluding outliers using the Mahanalobis distance. We then extend the statistic to multi-level designs, and propose a new statistical test termed \(ANOVA^2_{circ}\). This has identical assumptions to \(T^2_{circ}\), and is shown to be more sensitive than MANOVA when these assumptions are met. We demonstrate how these tests might be applied to two publicly available empirical data sets, and suggest concrete guidance for choosing which test to run. Implementations of these novel tools are provided as \emph{R} functions, and we hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.

@@

More recently, Victor and Mast (1991) proposed a simpler version of \(T^2\), called \(T^2_{circ}\). The \(T^2_{circ}\) statistic makes the strong assumption that the dependent variables (real and imaginary values) are uncorrelated and have equal variance. When these conditions are met, the test statistic is defined as:

@@

where \(x_j\) denotes the \(j\)th observation of the dependent variables, and all other terms retain their previous meanings. Notice that no covariance term is present in equation \eqref{eq:t2c}, because of the independence assumption. This makes the statistic simpler to calculate, but causes problems when the assumption is violated.

@@

All scripts used to generate this manuscript are available at: \url{https://github.com/bakerdh/FourierStats}. These include \emph{R} functions to implement one-sample Hotelling's (1931) \(T^2\) statistic, Victor \& Mast's (1991) \(T^2_{circ}\) statistic, and the condition index and \(ANOVA^2_{circ}\) statistics proposed in this paper.

@@

The upper row of Figure \ref{fig:amphists} shows scatterplots of simulated Fourier components, expressed using real (x) and imaginary (y) components. The amplitudes are the lengths of the lines joining each grey point to the origin. The lower row in Figure \ref{fig:amphists} shows distributions of amplitudes for the same set of signal strengths. These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's d \textgreater{} 2; Cohen's d is the mean difference scaled by the standard deviation, see Cohen (1988)). One consequence of this is that T-tests will potentially have an inflated Type 1 error (false positive) rate for many signals encountered empirically.

@@

Here we replicate and extend these simulations (see Figure \ref{fig:powerfig}), and confirm that the regime where \(T^2_{circ}\) has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \ref{fig:powerfig}f). However, for effect sizes around 0.5 \textless{} \emph{d} \textless{} 1, \(T^2_{circ}\) is more sensitive even with around 16 observations. This advantage is lost for large sample sizes (N \textgreater{} 32) and large effect sizes (d \textgreater{} 2 when N \textgreater{} 8). These simulations suggest a straightforward heuristic - there is no advantage to using the \(T^2_{circ}\) statistic for sample sizes, so its use should be restricted to small sample studies.

@@

Although \(T^2_{circ}\) can be more sensitive than \(T^2\), this greater sensitivity relies on meeting the test's more stringent assumptions. The two variables must be independent, and of equal variance. These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not. Figure \ref{fig:falsealarms} shows the results of simulations with randomly generated bivariate data in which no signal is present. When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of \(\alpha = 0.05\) (dashed line). However, as the data become increasingly correlated (Figure \ref{fig:falsealarms}a), or the variances of the two dependent variables more disparate (Figure \ref{fig:falsealarms}b), the Type I error rate of the \(T^2_{circ}\) statistic (shown in red) increases by almost a factor of 2. In contrast, the \(T^2\) statistic, which explicitly takes account of the covariance matrix (see equation \eqref{eq:t2eq}) shows no increase (black curves).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/falsealarms-1.pdf}
\caption{\label{fig:falsealarms}Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b). Estimates are for 100000 simulations per condition, with N=10 observations. The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.}
\end{figure}

@@

@@

Following \({ANOVA}^2_{circ}\), one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that standard Bonferroni corrections are applied. Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary). A repeated measures version of \({ANOVA}^2_{circ}\) has also been implemented following the same logic. The project repository contains \emph{R} scripts for both of these functions.

@@

The flowchart in Figure \ref{fig:flowchart} contains a proposed decision structure for the analysis of periodic data. Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios. Comparisons with one or two conditions should be tested with the \(T^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, and the \(T^2\) statistic if not. Comparisons with more than two conditions should be tested with the \(ANOVA^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not. Some MANOVA implementations cannot deal with random factors (repeated measures) in complex factorial designs. In \emph{R}, the \emph{Manova} function from the \emph{car} package appears to be capable of the necessary calculations, as does the commercial software package SPSS.

@@

Figure \ref{fig:mousedata}a shows Fourier amplitude spectra at two frontal electrodes (marked black in the insets) for auditory stimulation (red) and optogenetic stimulation (blue), each at 40 Hz. There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality. Indeed, a paired t-test on the amplitudes reveals no significant difference (t=0.84, df=5, p=0.44). However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \ref{fig:mousedata}b). The condition index test was non-significant for both conditions (sound: CI = 1.59, \emph{p} = 0.66; light: CI = 1.69, \emph{p} = 0.59), so a paired-samples \(T^2_{circ}\) test was conducted. This revealed a significant difference between conditions (\(T^2_{circ}\) = 1.39, \(F_{(2,10)}\) = 8.32, \emph{p} = 0.007) with an effect size of D = 1.02. This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al. (2019) went on to explore interactions between these two signals.

@@

@@

Vilidaite et al. (2018) measured visual responses to flickering grating stimuli in a large sample of 100 adults. Each participant completed a series of 11-second trials, in which stimuli of different contrasts flickered at 7Hz (on-off sinusoidal flicker). Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \ref{fig:humanSSVEP}), were well-isolated in the Fourier domain, and increased with stimulus contrast. For this analysis, we took responses from electrode \emph{Oz} at the occipital pole (black points in the upper row of Figure \ref{fig:humanSSVEP}), averaged across repetition for each participant. Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \ref{fig:humanSSVEP}. Any participant that contributed at least one outlier was excluded, leaving a total of 89 participants for the main analysis.

@@

\caption{Summary of human SSVEP data. Upper row shows scalp distributions of Fourier amplitudes at 7Hz for stimuli of increasing contrasts (blue shading indicates higher amplitudes). Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (black point). Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the centroids.}\label{fig:humanSSVEP}
\end{figure}

With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = 1.2, all \emph{p} \textgreater{} 0.19). A repeated measures \(ANOVA^2_{circ}\) test was conducted, revealing a significant effect of stimulus contrast (\(F_{(12,1056)}\) = 38.9, p \textless{} 0.001). Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.08\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.4, D = 0.13, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.3, D = 0.19, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.5, D = 0.14, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.34, \(F_{(2,176)}\) = 35.3, D = 0.33, \emph{p} \textless{} 0.001). The study by Vilidaite et al. (2018) compared SSVEP responses between individuals with and without autism, as well as in a \emph{Drosophila} genetic model of developmental disorders. The raw data are available at: \url{http://dx.doi.org/10.17605/OSF.IO/Y4N5K}.

@@

It is worth pointing out that the statistical tests discussed in this paper are applicable only when the signal phase is expected to be consistent across observations. This is the case for most paradigms in which the nervous system is driven by a periodic stimulus. However, they are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms (Berger 1929; Buzsáki and Draguhn 2004), which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals (e.g. Canolty and Knight 2010). When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average (Baker et al. 2020). This practice is also necessary in order to use the multivariate methods discussed here, as the alternative is to discard the phase information and average amplitudes instead, rendering the data univariate.

The present paper has focussed on the Frequentist statistical tradition. However there are many advantages to the Bayesian approach, in which one can make direct quantitative comparisons of the evidence supporting both the experimental and null hypotheses. Subject to determining appropriate priors, Bayes factor scores might be calculated for all of the statistics considered here, much as has been done previously for univariate T-tests (Rouder et al. 2009). However this is a non-trivial undertaking, and is beyond the scope of the current paper.

Another possibility is to use machine learning techniques such as multivariate pattern analysis (MVPA) to analyse periodic data. This involvess training a classifier algorithm to distinguish between two (or more) experimental conditions or states, and then assessing classifier accuracy on fresh data. If different conditions produce distinct patterns of neural response, then classifier accuracy will be above chance. Such methods have been hugely influential in the fMRI literature, and for analysing event-related potential data collected using EEG or MEG. However, they have not been widely applied to steady-state data (though see West et al. (2015) for an example). In principle, the real and imaginary Fourier components can be treated as separate dependent variables, along with different recording locations and/or frequencies. This approach has the potential to offer sensitive, high-powered statistical tests that circumvent many of the shortcomings associated with parametric statistics.

Even when statistics are conducted using both the real and imaginary Fourier components, it is still typical to visualise the mean amplitudes. Several approaches to calculating appropriate error bars have been proposed. For example, Pei et al. (2017) suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude. This approach is somewhat computationally demanding, though we have implemented an \emph{R} function to calculate error bars using this method. However, we feel that bootstrap resampling offers a powerful and general alternative for calculating 95\% confidence intervals on amplitudes. This is achieved by resampling the complex data (with replacement) and calculating a resampled complex mean. The amplitude is then derived for this resampled mean, and the procedure repeated a large number of times (1000 or 10000 repetitions is typical) to build up a population of resampled amplitudes. Upper and lower 95\% confidence intervals on the amplitude can then be taken at appropriate quantiles of this population.

@@

\leavevmode\hypertarget{ref-Baker2020}{}%
Baker, D.H., G. Vilidaite, F.A. Lygo, A.K. Smith, T.R. Flack, A.D. Gouws, and T.J. Andrews. 2020. ``Power Contours: Optimising Sample Size and Precision in Experimental Psychology and Human Neuroscience.'' \emph{Psychological Methods}. \url{https://doi.org/10.1037/met0000337}.

@@

\leavevmode\hypertarget{ref-Gray2020}{}%
Gray, Katie L H, Tessa R Flack, Miaomiao Yu, Freya A Lygo, and Daniel H Baker. 2020. ``Nonlinear Transduction of Emotional Facial Expression.'' \emph{Vision Res} 170 (May): 1--11. \url{https://doi.org/10.1016/j.visres.2020.03.004}.

@@

\leavevmode\hypertarget{ref-Liu-Shuang2014}{}%
Liu-Shuang, Joan, Anthony M Norcia, and Bruno Rossion. 2014. ``An Objective Index of Individual Face Discrimination in the Right Occipito-Temporal Cortex by Means of Fast Periodic Oddball Stimulation.'' \emph{Neuropsychologia} 52 (January): 57--72. \url{https://doi.org/10.1016/j.neuropsychologia.2013.10.022}.

@@

@@

\leavevmode\hypertarget{ref-Pei2017}{}%
Pei, Francesca, Stefano Baldassi, Jeffrey J Tsai, Holly E Gerhard, and Anthony M Norcia. 2017. ``Development of Contrast Normalization Mechanisms During Childhood and Adolescence.'' \emph{Vision Res} 133 (April): 12--20. \url{https://doi.org/10.1016/j.visres.2016.03.010}.

@@

@@

@@

\leavevmode\hypertarget{ref-West2015}{}%
West, Ryan J H, Christopher J H Elliott, and Alex R Wade. 2015. ``Classification of Parkinson's Disease Genotypes in Drosophila Using Spatiotemporal Profiling of Vision.'' \emph{Sci Rep} 5 (November): 16933. \url{https://doi.org/10.1038/srep16933}.

>>>>

\title{Statistical analysis of periodic data in neuroscience}
\author{Daniel H. Baker}
\date{2020-12-30}

@@

Many experimental paradigms in neuroscience involve driving the nervous system with periodic sensory stimuli. Neural signals recorded with a variety of techniques will then include phase-locked oscillations at the stimulation frequency. The analysis of such data often involves standard univariate statistics such as T-tests, conducted on the Fourier amplitude components. However, the assumptions of these tests will often be violated because amplitudes are not normally distributed, and furthermore weak signals might be missed if the phase information is discarded. An alternative approach is to conduct multivariate statistical tests using the real and imaginary Fourier components. Here we compare the performance of two multivariate extensions of the T-test: Hotelling's \(T^2\) and a variant called \(T^2_{circ}\). We develop a novel test of the assumptions of \(T^2_{circ}\) based on the condition index, and suggest a heuristic for excluding outliers using the Mahanalobis distance. We then extend the statistic to multi-level designs, and propose a new statistical test termed \(ANOVA^2_{circ}\). This has identical assumptions to \(T^2_{circ}\), and is shown to be more sensitive than MANOVA when these assumptions are met. We demonstrate how these tests might be applied to two publicly available empirical data sets, and suggest concrete guidance for choosing which test to run. Implementations of these novel tools are provided as \emph{R} functions, and we hope that their wider adoption will improve the sensitivity of statistical inferences involving periodic data.

@@

More recently, Victor and Mast (1991) proposed a simpler version of \(T^2\), called \(T^2_{circ}\). The \(T^2_{circ}\) statistic makes the strong assumption that the dependent variables (real and imaginary values) are uncorrelated and have equal variance. When these conditions are met, the test statistic for the one-sample case is defined as:

@@

where \(x_j\) denotes the \(j\)th observation of the dependent variables, and all other terms retain their previous meanings. Notice that no covariance term is present in equation \eqref{eq:t2c}, because of the independence assumption. This makes the statistic simpler to calculate, but causes problems when the assumption is violated (as we will demonstrate below). Conceptually, this statistic takes the vector difference between the centroid (\(\bar{x}\)) and a comparison point (\(\mu\)), and scales by the lengths of the residual vector lines joining each data point to the centroid. Two-sample and repeated measures versions of the \(T^2_{circ}\) statistic are also possible.

@@

All scripts used to generate this manuscript are available at: \url{https://github.com/bakerdh/FourierStats}. This includes an \emph{R} package, featuring functions to implement one-sample Hotelling's (1931) \(T^2\) statistic, Victor \& Mast's (1991) \(T^2_{circ}\) statistic, and the condition index and \(ANOVA^2_{circ}\) statistics proposed in this paper.

@@

The upper row of Figure \ref{fig:amphists} shows scatterplots of simulated Fourier components, expressed using real (\emph{x}) and imaginary (\emph{y}) components. The amplitudes are the lengths of the lines joining each grey point to the origin. The lower row in Figure \ref{fig:amphists} shows distributions of amplitudes for the same set of signal strengths. These distributions only approach normality when the signal strength is more than twice the standard deviation (Cohen's \emph{d} \textgreater{} 2; Cohen's \emph{d} is the mean difference scaled by the standard deviation, see Cohen (1988)). One consequence of this is that T-tests will potentially have an inflated Type 1 error (false positive) rate for many signals encountered empirically.

@@

Here we replicate and extend these simulations (see Figure \ref{fig:powerfig}), and confirm that the regime where \(T^2_{circ}\) has greater power occurs particularly for large effect sizes and small sample sizes (see Figure \ref{fig:powerfig}f). However, for effect sizes around 0.5 \textless{} \emph{d} \textless{} 1, \(T^2_{circ}\) is more sensitive even with around 16 observations. This advantage is lost for large sample sizes (N \textgreater{} 32) and large effect sizes (\emph{d} \textgreater{} 2 when N \textgreater{} 8). These simulations suggest a straightforward heuristic - there is no advantage to using the \(T^2_{circ}\) statistic for large sample sizes (when N \textgreater{} 32), so its use should be restricted to small sample studies.

@@

Although \(T^2_{circ}\) can be more sensitive than \(T^2\), this greater sensitivity relies on meeting the test's more stringent assumptions. The two variables must be independent (i.e.~uncorrelated), and of equal variance. These restrictions may hold for some data sets, but it is instructive to ask what happens when they do not. Figure \ref{fig:falsealarms} shows the results of simulations with randomly generated bivariate data in which no signal is present. When the data are uncorrelated and have equal variance (mid-points of the functions in each panel), both tests have the nominal Type I error (false positive) rate of \(\alpha = 0.05\) (dashed line). However, as the data become increasingly correlated (Figure \ref{fig:falsealarms}a), or the variances of the two dependent variables more disparate (Figure \ref{fig:falsealarms}b), the Type I error rate of the \(T^2_{circ}\) statistic (shown in red) increases by almost a factor of 2. In contrast, the \(T^2\) statistic, which explicitly takes account of the covariance matrix (see equation \eqref{eq:t2eq}) shows no increase (black curves).

\begin{figure}
\centering
\includegraphics{manuscript_files/figure-latex/falsealarms-1.pdf}
\caption{\label{fig:falsealarms}Simulations showing the Type I error rate for both tests as a function of the correlation between two variables (a) and the ratio of variances (b). Estimates are for 100000 simulations per condition, with N = 10 observations. The icons at the foot of each panel show example scatterplots with bounding ellipses and eigenvectors.}
\end{figure}

@@

\hypertarget{controlling-for-multiple-comparisons-across-location-and-time}{%
\section{Controlling for multiple comparisons across location and time}\label{controlling-for-multiple-comparisons-across-location-and-time}}

In many studies, it is important to compare responses over space and/or time. However with large numbers of sensors, voxels or temporal epochs, the familywise error rate quickly becomes problematic, inflating the Type I error (false positive) rate. Solutions such as Bonferroni correction, which adjust the alpha level based on the number of comparisons, are overly conservative and can obscure real effects by dramatically reducing power. An alternative approach is to use cluster correction methods to control the Type I error rate, using mass univariate tests (e.g. Maris and Oostenveld 2007). These typically involve summing test statistics such as T-values across adjacent significant locations and/or moments in time. The summed test statistic is compared to a null distribution generated from the same data by randomly permuting condition labels (or the sign of the data for one-sample tests). Such methods control the Type I error rate without substantially reducing statistical power. The same approach can be straightforwardly applied to the \(T^2\) and \(T^2_{circ}\) statistics. This allows a principled approach to identifying clusters of significant sensors, timepoints or frequencies responding to periodic stimuli. The \emph{FourierStats} package includes an implementation of this method with options for multivariate statistics (the \emph{clustercorrect} function).

@@

Following \({ANOVA}^2_{circ}\), one could calculate \(T^2_{circ}\) statistics to make post-hoc pairwise comparisons between conditions, providing that standard multiple comparison corrections are applied (e.g.~Bonferroni correction). Univariate ANOVAs on the real and imaginary components are unlikely to be informative, as the relative magnitudes depend on stimulus phase (which is arbitrary). A repeated measures version of \({ANOVA}^2_{circ}\) has also been implemented following the same logic. The project repository contains \emph{R} scripts for both of these functions.

@@

The flowchart in Figure \ref{fig:flowchart} contains a proposed decision structure for the analysis of periodic data, once any outliers have been removed. Initially, data for each condition should be tested against the expected distribution of eigenvalue ratios. Comparisons with one or two conditions should be tested with the \(T^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, and the \(T^2\) statistic if not. Comparisons with more than two conditions should be tested with the \(ANOVA^2_{circ}\) statistic if the eigenvalue ratios are consistent with circularity, or a MANOVA if not. Some MANOVA implementations cannot deal with random factors (repeated measures) in complex factorial designs. In \emph{R}, the \emph{Manova} function from the \emph{car} package appears to be capable of the necessary calculations, as does the commercial software package SPSS.

@@

@@

Figure \ref{fig:mousedata}a shows Fourier amplitude spectra at two frontal electrodes (marked black in the insets) for auditory stimulation (red) and optogenetic stimulation (blue), each at 40 Hz. There is a clear frequency-locked signal with approximately equal amplitude for each stimulation modality. Indeed, a paired univariate T-test on the amplitudes reveals no significant difference (\emph{t} = 0.84, \emph{df} = 5, \emph{p} = 0.44). However, inspection of the complex Fourier components for each condition suggests evidence of a phase difference between the two modalities (see Figure \ref{fig:mousedata}b). The condition index test was non-significant for both conditions (sound: CI = 1.59, \emph{p} = 0.66; light: CI = 1.69, \emph{p} = 0.59), so a paired-samples \(T^2_{circ}\) test was conducted. This revealed a significant difference between conditions (\(T^2_{circ}\) = 1.39, \(F_{(2,10)}\) = 8.32, \emph{p} = 0.007) with an effect size of D = 1.02. This demonstrates that both sound and (optogenetic) light are able to entrain neural responses - the original study by Hwang et al. (2019) went on to explore interactions between these two signals.

@@

Vilidaite et al. (2018) measured visual responses to flickering grating stimuli in a large sample of 100 adults. Each participant completed a series of 11-second trials, in which stimuli of different contrasts flickered at 7Hz (on-off sinusoidal flicker). Responses were strongest at occipital electrodes over visual cortex (see upper row of Figure \ref{fig:humanSSVEP}), were well-isolated in the Fourier domain, and increased with stimulus contrast. Significant activity was evident at 4\% contrast and above following cluster correction (with very stringent alpha levels given the high power of this data set), as indicated by the red electrodes in the upper row of \ref{fig:humanSSVEP}. For the main analysis, we took responses from electrode \emph{Oz} at the occipital pole (black points in the upper row of Figure \ref{fig:humanSSVEP}), averaged across repetition for each participant. Each condition included some outlier points with Mahalanobis distances exceeding 3, marked red in the lower row of Figure \ref{fig:humanSSVEP}. Any participant that contributed at least one outlier was excluded, leaving a total of 89 participants for the main analysis.

@@

\caption{Summary of human SSVEP data. Upper row shows scalp distributions of Fourier amplitudes at 7Hz for stimuli of increasing contrasts (blue shading indicates higher amplitudes). Electrodes marked in red indicate cluster-corrected significance.  Lower row shows scatterplots of complex (x = real, y = imaginary) Fourier components for 100 participants per condition, from electrode Oz (black point). Red points are outliers with Mahalanobis distances exceeding 3, and blue points mark the centroids.}\label{fig:humanSSVEP}
\end{figure}

With the outlier points removed, all seven conditions resulted in non-significant condition index tests (largest CI = 1.20, all \emph{p} \textgreater{} 0.19). A repeated measures \(ANOVA^2_{circ}\) test was conducted, revealing a significant effect of stimulus contrast (\(F_{(12,1056)}\) = 38.9, \emph{p} \textless{} 0.001). Pairwise \(T^2_{circ}\) statistics comparing the baseline (0\% contrast) condition to each subsequent condition (Bonferroni corrected for 6 tests to \(\alpha = 0.08\)) revealed significant differences at 8\% contrast (\(T^2_{circ}\) = 0.32, \(F_{(2,176)}\) = 28.43, D = 0.13, \emph{p} \textless{} 0.001), 16\% contrast (\(T^2_{circ}\) = 0.28, \(F_{(2,176)}\) = 25.25, D = 0.19, \emph{p} \textless{} 0.001), 32\% contrast (\(T^2_{circ}\) = 0.10, \(F_{(2,176)}\) = 8.55, D = 0.14, \emph{p} \textless{} 0.001) and 64\% contrast (\(T^2_{circ}\) = 0.40, \(F_{(2,176)}\) = 35.35, D = 0.33, \emph{p} \textless{} 0.001). The study by Vilidaite et al. (2018) compared SSVEP responses between individuals with and without autism, as well as in a \emph{Drosophila} genetic model of developmental disorders. The raw data are available at: \url{http://dx.doi.org/10.17605/OSF.IO/Y4N5K}.

@@

It is worth pointing out that the statistical tests discussed in this paper are applicable only when the signal phase is expected to be consistent across observations. This is the case for most paradigms in which the nervous system is driven by a periodic stimulus. However, they are less obviously applicable to the analysis of endogenous neural oscillations and brain rhythms (Berger 1929; Buzsáki and Draguhn 2004), which will typically have random phase and broader bandwidths in the Fourier domain: other analysis methods have been developed for such signals (e.g. Canolty and Knight 2010). When phases are consistent across repetitions, greater statistical power can be obtained by coherently averaging across repetitions to obtain a participant-level average (Baker et al. 2021). This practice is also necessary in order to use the multivariate methods discussed here, as the alternative is to discard the phase information and average amplitudes instead, rendering the data univariate.

The present paper has focussed on the Frequentist statistical tradition. However there are many advantages to the Bayesian approach, in which one can make direct quantitative comparisons of the evidence supporting both the experimental and null hypotheses (Jeffreys 1961). Subject to determining appropriate priors, Bayes factor scores might be calculated for all of the statistics considered here, much as has been done previously for univariate T-tests (Rouder et al. 2009) and ANOVA (Rouder et al. 2017). However this is a non-trivial undertaking, and is beyond the scope of the current paper.

Another possibility is to use machine learning techniques such as multivariate pattern analysis (MVPA) to analyse periodic data. This involvess training a classifier algorithm to distinguish between two (or more) experimental conditions or states, and then assessing classifier accuracy on fresh data. If different conditions produce distinct patterns of neural response, then classifier accuracy will be above chance. Such methods have been hugely influential in the fMRI literature (Schwarzkopf and Rees 2011), and for analysing event-related potential data collected using EEG or MEG (Grootswagers, Wardle, and Carlson 2017). However, they have not been widely applied to steady-state data (though see West et al. (2015) for an example). In principle, the real and imaginary Fourier components can be treated as separate dependent variables, along with different recording locations and/or frequencies. This approach has the potential to offer sensitive, high-powered statistical tests that circumvent many of the shortcomings associated with traditional statistics.

Even when statistics are conducted using both the real and imaginary Fourier components, it is still typical to visualise the mean amplitudes. Several approaches to calculating appropriate error bars have been proposed. For example, Pei et al. (2017) suggest calculating the nearest and farthest points from the origin on the bounding ellipse, and using these to derive standard errors for the amplitude. This approach is somewhat demanding, though we have implemented an \emph{R} function to calculate error bars using this method (available through the \emph{getampCI} function). However, we feel that bootstrap resampling offers a powerful and general alternative for calculating 95\% confidence intervals on amplitudes. This is achieved by resampling the complex data (with replacement) and calculating a resampled complex mean. The amplitude is then derived for this resampled mean, and the procedure repeated a large number of times (1000 or 10000 repetitions is typical) to build up a population of resampled amplitudes. Upper and lower 95\% confidence intervals on the amplitude can then be taken at appropriate quantiles of this population.

@@

\leavevmode\hypertarget{ref-Baker2021}{}%
Baker, D.H., G. Vilidaite, F.A. Lygo, A.K. Smith, T.R. Flack, A.D. Gouws, and T.J. Andrews. 2021. ``Power Contours: Optimising Sample Size and Precision in Experimental Psychology and Human Neuroscience.'' \emph{Psychological Methods}. \url{https://doi.org/10.1037/met0000337}.

@@

\leavevmode\hypertarget{ref-Gray2020}{}%
Gray, Katie L H, Tessa R Flack, Miaomiao Yu, Freya A Lygo, and Daniel H Baker. 2020. ``Nonlinear Transduction of Emotional Facial Expression.'' \emph{Vision Res} 170: 1--11. \url{https://doi.org/10.1016/j.visres.2020.03.004}.

\leavevmode\hypertarget{ref-Grootswagers2017}{}%
Grootswagers, Tijl, Susan G Wardle, and Thomas A Carlson. 2017. ``Decoding Dynamic Brain Patterns from Evoked Responses: A Tutorial on Multivariate Pattern Analysis Applied to Time Series Neuroimaging Data.'' \emph{J Cogn Neurosci} 29 (4): 677--97. \url{https://doi.org/10.1162/jocn_a_01068}.

@@

\leavevmode\hypertarget{ref-Jeffreys1961}{}%
Jeffreys, H. 1961. \emph{Theory of Probability}. 3rd ed. Oxford University Press.

\leavevmode\hypertarget{ref-Liu-Shuang2014}{}%
Liu-Shuang, Joan, Anthony M Norcia, and Bruno Rossion. 2014. ``An Objective Index of Individual Face Discrimination in the Right Occipito-Temporal Cortex by Means of Fast Periodic Oddball Stimulation.'' \emph{Neuropsychologia} 52: 57--72. \url{https://doi.org/10.1016/j.neuropsychologia.2013.10.022}.

@@

\leavevmode\hypertarget{ref-Maris2007}{}%
Maris, Eric, and Robert Oostenveld. 2007. ``Nonparametric Statistical Testing of EEG- and MEG-Data.'' \emph{J Neurosci Methods} 164 (1): 177--90. \url{https://doi.org/10.1016/j.jneumeth.2007.03.024}.

@@

\leavevmode\hypertarget{ref-Pei2017}{}%
Pei, Francesca, Stefano Baldassi, Jeffrey J Tsai, Holly E Gerhard, and Anthony M Norcia. 2017. ``Development of Contrast Normalization Mechanisms During Childhood and Adolescence.'' \emph{Vision Res} 133: 12--20. \url{https://doi.org/10.1016/j.visres.2016.03.010}.

@@

\leavevmode\hypertarget{ref-Rouder2017}{}%
Rouder, Jeffrey N, Richard D Morey, Josine Verhagen, April R Swagman, and Eric-Jan Wagenmakers. 2017. ``Bayesian Analysis of Factorial Designs.'' \emph{Psychol Methods} 22 (2): 304--21. \url{https://doi.org/10.1037/met0000057}.

@@

\leavevmode\hypertarget{ref-Schwarzkopf2011}{}%
Schwarzkopf, D S, and G Rees. 2011. ``Pattern Classification Using Functional Magnetic Resonance Imaging.'' \emph{Wiley Interdiscip Rev Cogn Sci} 2 (5): 568--79. \url{https://doi.org/10.1002/wcs.141}.

@@

\leavevmode\hypertarget{ref-West2015}{}%
West, Ryan J H, Christopher J H Elliott, and Alex R Wade. 2015. ``Classification of Parkinson's Disease Genotypes in Drosophila Using Spatiotemporal Profiling of Vision.'' \emph{Sci Rep} 5: 16933. \url{https://doi.org/10.1038/srep16933}.
